---
title: "Exploratory data analysis- human activity"
author: "May Phang"
date: "November 8, 2016"
output: ioslides_presentation
---

```{r setup, include=FALSE, fig.align = 'center'}
knitr::opts_chunk$set(echo = FALSE)
```

```{r, data, echo = FALSE, cache = TRUE}
# read in necessary libraries
suppressMessages(library(dplyr))
library(reshape)

setwd("C:/Users/ymp/Documents/Kaggle/201611_UCIHumanActivity")
############## Step 1: Merges the training and the test sets to create one data set ##############

## download zip if it doesnt exist ##
if(!file.exists("getdata-projectfiles-UCI HAR Dataset.zip")|!file.exists("UCI HAR Dataset")){
  ZipURL <- "https://d396qusza40orc.cloudfront.net/getdata%2Fprojectfiles%2FUCI%20HAR%20Dataset.zip"
  download.file(ZipURL, destfile="./getdata-projectfiles-UCI HAR Dataset.zip")
  ZipFile <- unzip("getdata-projectfiles-UCI HAR Dataset.zip", exdir = getwd())
}


## create tables from the unzipped files ##
dir_name <- "./UCI HAR Dataset/"
features_data <- read.table(paste(dir_name, "features.txt", sep="")) # read in features name, these are variables for each test and training datasets
features <- features_data$V2 #these are the column names for each dataset
rm("features_data") # clear unnecessary data from the workspace for memory management

# read in the descriptive activity names data
Desc_Activity <- read.table("./UCI HAR Dataset/activity_labels.txt", col.names = c("activityID", "ActivityDesc"))

## read in training data ##
sub_train <- read.table(paste(dir_name, "train/subject_train.txt", sep=""), col.names = "SubjectID") #corresponding subject ID
X_train <- read.table(paste(dir_name, "train/X_train.txt", sep=""), col.names = features) # training features
Y_train <- read.table(paste(dir_name, "train/Y_train.txt", sep=""), col.names = "activityID") # these are the activity labels (as ID)
train  <- cbind(X_train, sub_train, Y_train) # combine the features and labels into 1 training dataset

train <- merge(train, Desc_Activity)

# fix up the names
names(train) <- gsub("(\\.{1,})", ".", names(train))
# get rid of the repeated words for Body
names(train) <- tolower(gsub("BodyBody", "Body", names(train)))
names(train) <- tolower(names(train))
train$subjectid <- as.factor(train$subjectid)
train$activitydesc <- as.factor(train$activitydesc)


## read in test data ##
sub_test <- read.table(paste(dir_name, "test/subject_test.txt", sep=""), col.names = "SubjectID") #corresponding subject ID
X_test <- read.table(paste(dir_name, "test/X_test.txt", sep=""), col.names = features) # testing features
Y_test <- read.table(paste(dir_name, "test/Y_test.txt", sep=""), col.names = "activityID") # these are the activity labels (as ID)
test  <- cbind(X_test, sub_test, Y_test) # combine the features and labels into 1 testing dataset

test <- merge(test, Desc_Activity)

# fix up the names
names(test) <- gsub("(\\.{1,})", ".", names(test))
# get rid of the repeated words for Body
names(test) <- tolower(gsub("BodyBody", "Body", names(test)))
names(test) <- tolower(names(test))
test$subjectid <- as.factor(test$subjectid)
test$activitydesc <- as.factor(test$activitydesc)

# find the columns that contain the relevant words
Col_Mean <- names(select(train, contains("mean", ignore.case = FALSE))) # dont want to get the angles
Col_Std <- names(select(train, contains("std", ignore.case = "FALSE"))) # dont want to get the angles
# Col_Retain <- c("SubjectID", "activityID", Col_Mean, Col_Std)# create a vector of the columns to retain

```

## Introduction

This short slide deck includes interactive visualisations of exploratory data analysis carried out on the UCI Machine Learning data on Human Activity recognition. 

These includes: 

- Response variable distribution
- fbodyacc s.d. 3D visualisation
- Metric box plots by subject ID (walking)

## Response variable distribution

```{r dist, cache = TRUE, echo = FALSE, fig.align = 'center'}
suppressMessages(library(plotly))
suppressMessages(library(ggplot2))
suppressMessages(library(mlr))

activity <- sort(unique(train$activitydesc))
subjectid.cnt <- table(train$subjectid, train$activitydesc)
# colnames(subjectid.cnt) == activity
m <- list(  l = 50,
            r = 20,
            b = 150,
            t = 25,
            pad = 4)
plot_ly(data = train, x = ~ activity, y = ~subjectid.cnt[1, ], name = row.names(subjectid.cnt)[1], type = "bar") %>% 
  add_trace(y = ~subjectid.cnt[2, ], name = row.names(subjectid.cnt)[2]) %>%
  add_trace(y = ~subjectid.cnt[3, ], name = row.names(subjectid.cnt)[3]) %>%
  add_trace(y = ~subjectid.cnt[4, ], name = row.names(subjectid.cnt)[4]) %>%
  add_trace(y = ~subjectid.cnt[5, ], name = row.names(subjectid.cnt)[5]) %>%
  add_trace(y = ~subjectid.cnt[6, ], name = row.names(subjectid.cnt)[6]) %>%
  add_trace(y = ~subjectid.cnt[7, ], name = row.names(subjectid.cnt)[7]) %>%
  add_trace(y = ~subjectid.cnt[8, ], name = row.names(subjectid.cnt)[8]) %>%
  add_trace(y = ~subjectid.cnt[9, ], name = row.names(subjectid.cnt)[9]) %>%
  add_trace(y = ~subjectid.cnt[10, ], name = row.names(subjectid.cnt)[10]) %>%
  add_trace(y = ~subjectid.cnt[11, ], name = row.names(subjectid.cnt)[11]) %>%
  add_trace(y = ~subjectid.cnt[12, ], name = row.names(subjectid.cnt)[12]) %>%
  add_trace(y = ~subjectid.cnt[13, ], name = row.names(subjectid.cnt)[13]) %>%
  add_trace(y = ~subjectid.cnt[14, ], name = row.names(subjectid.cnt)[14]) %>%
  add_trace(y = ~subjectid.cnt[15, ], name = row.names(subjectid.cnt)[15]) %>%
  add_trace(y = ~subjectid.cnt[16, ], name = row.names(subjectid.cnt)[16]) %>%
  add_trace(y = ~subjectid.cnt[17, ], name = row.names(subjectid.cnt)[17]) %>%
  add_trace(y = ~subjectid.cnt[18, ], name = row.names(subjectid.cnt)[18]) %>%
  add_trace(y = ~subjectid.cnt[19, ], name = row.names(subjectid.cnt)[19]) %>%
  add_trace(y = ~subjectid.cnt[20, ], name = row.names(subjectid.cnt)[20]) %>%
  add_trace(y = ~subjectid.cnt[21, ], name = row.names(subjectid.cnt)[21]) %>%
  # layout(autosize = F, width = 500, height = 650, margin = m, showlegend = F,
  layout(autosize = F, width = 800, height = 500, margin = m, showlegend = F,
         title = "Distribution of activity observations by subject ID", yaxis = list(title = "Observations"), 
         barmode = "stack", xaxis = list(tickangle = 90, textfont = list(size = 2.5), title = ""))

```

## fbodyacc s.d. 3D visualisation

```{r met3d, echo = FALSE, cache = TRUE}

plot_ly(data = train, x = train$fbodyacc.std.x, y = train$fbodyacc.std.y, z = train$fbodyacc.std.z, type = "scatter3d", 
        mode = "markers", color = train$activitydesc, marker = list(size = 4))
```


## Metric box plots by subject ID (walking)


```{r subid, echo = FALSE, cache = TRUE, warning = FALSE, message = FALSE}
plot_ly(data = subset(train, activitydesc = "WALKING"), y = train$fbodyacc.std.x, color = train$subjectid, type = "box") %>%
  layout(autosize = F, width = 800, height = 500)
# may imply need to form some sort of calibration so that the algorithm takes it into account (different heights)

```


